Contextual recruitment of selective attention can be updated via changes in task-relevance

### 2019-09-16 manuscript_pretty.Rmd
revision in progress

Real-world environments can offer a plethora of context cues. For instance, consider someone writing in a coffee shop. The coffee shop is a complex environment with numerous context cues (e.g., pictures on the wall, chairs, coffee, laptop), each of which we have different experiences with and afford different learned associations. Although a laptop might cue attentional processing helpful for writing, other features of the coffee shop might cue attentional processing related to other experiences like socializing and people-watching. Given that all of these cues are available to set attentional priorities, what determines which associative relationships will be used to guide attention? On the one hand, context-based retrieval of attentional priorities might be completely obligatory: the mere presence of learned cues might trigger attention in a ballistic manner [e.g., @schneider_controlled_1977; @shiffrin_controlled_1977]. A writer's attention in this case would be at the mercy of the environment. On the other hand, contextual cuing of attention may not be entirely obligatory. Our current task-goals, for instance, might constrain the expression of previously learned associations, allowing, for instance, the laptop to take priority as a contextual cue over other cues that are irrelevant to the task at hand.

Historically, perspectives of automaticity and attentional control have emphasized the role of practice and likened the learning process to the development of automaticity in skill-learning [@schneider_controlled_1977; @shiffrin_controlled_1977; @logan_toward_1988; @logan_attention_1992]. Under this view, forming an association requires many repeated experiences pairing a cue with a set of attentional priorities. Once formed however, it becomes habitual or hard-wired and the mere presence of the cue is suffificient to automatically adjust attention. Learned associations are therefore ballistic, difficult to suppress, and difficult to unlearn [e.g., @schneider_controlled_1977; @shiffrin_controlled_1977; @posner_attention_1975; @posner_facilitation_1975; @treisman_contextual_1960; @treisman_selective_1964; @treisman_strategies_1969]. This view is reflected in modern work on contextual cuing where it is assumed that learned associations require practice [e.g., @lehle_fly_2008; @crump_learning_2016], form without awareness or intention, and are triggered automatically by the mere presence of the cue [@crump_context-specific_2006; @crump_contextual_2010; @crump_context-specific_2008; @king_priming_2012; @reuss_unconscious_2014; though, see @brosowsky_context-specific_2016].

Memory-based perspectives however, take a more nuanced view [@norman_memory_1969; @norman_memory_1976; @abrahamse_grounding_2016; @brosowsky_memory-guided_2018; @crump_learning_2016; @egner_creatures_2014]. Here, practice is only important to the extent that it improves memory retrieval. Learning an association is the process of encoding experiences, where attentional priorities and cues co-occur, into memory. Memory retrieval is a similarity-based, cue-driven process [e.g., @jacoby_interpreting_1978; @hintzman_minerva_1984; @hintzman_schema_1986; @hintzman_judgments_1988]. Presumably, the features of our immediate environment automatically cue memory and reinstate the retrieved attentional priorites [@logan_toward_1988; @logan_attention_1992]. If a few prior experiences are easily retrievable--memories are distinct or very similar to the retrieval cue--it could support the reinstatement of a learned attentional association despite very little practice [e.g., @brosowsky_memory-guided_2018]. Conversely, if prior experiences are difficult to retrieve--memories are in or low in similarity to the retrieval cue--then learned associations may not be expressed, even after much practice. To the extent that you could manipulate memory retrieval, you should be able manipulate the expression of learned associations.

@@

Recent work lends some credence to the possibility that task-relevance is necessary for establishing and using context-attention associations. The most commonly used context cue is location [@brosowsky_context-specific_2016; @corballis_independent_2003; @crump_context-specific_2006; @crump_context-specific_2008; @crump_learning_2016; @weidler_attentional_2018; @weidler_transfer_2016], often claimed to be irrelevant to the on-going conflict task. Although in some sense, the location is irrelevant for identifying the color of a word in a Stroop task, it seems disingenuous to consider it completely task-irrelevant. In order to identify the target, the participant must first identify and orient to its location on the screen. In this is sense, it is relevant to their ability to complete the task. Further, there is some evidence that location information may receive priority during encoding [@logan_what_1998; @mayr_spatial_1996]. Similar arguments could be made about other context features that are inherent to processing the target stimulus, like font-type [@bugg_multiple_2008] and color [@vietze_context_2009].

We are aware of only one study that has looked at task-relevance within the CSPC literature. Crump, Vaquero, & Milliken [-@crump_context-specific_2008] used a Stroop prime-probe task where the distractor word is presented first followed by a color patch probe. Critically, they used the shape of the color patch probe, either a square or a circle, as the contextual dimension; one associated with a high frequency of conflict, the other with a low frequency. In one experiment participants were made aware of the shape-proportion contingencies but given no additional instructions. In the second, participants were given a secondary task to count the number of trials that contained a square. The first experiment failed to find evidence for context-specificity, whereas the second did. Although not testing task-relevance directly, @canadas_social_2013 found similar effects. In this study they used images of male and female faces as context cues in a flanker task. When given instructions to think of the faces as members of the gender categories they found CSPC effects, but when given instructions to think of the faces as individuals they did not. These results suggest that making a context dimension relevant to the on-going task is important for producing CSPC effects.

@@

Previous work has shown that shifting between list-wide proportion congruent phases produced asymmetrical effects. Specifically, @abrahamse_attention_2013 found that shifting from a mostly-congruent to mostly-incongruent list produced a large change in the size of the congruency effect, whereas shifting from a mostly-incongruent to mostly-congruent list produced a relatively small change in congruency effects. The authors posited a learned attentional modulation in the mostly-incongruent block had carried over into the mostly-congruent block. In our study, we were primarily interested in whether (a) participants could use context cues despite the presence of competing, contradictory cues, and (b) whether participants could switch from the use of one context cue to another. We therefore randomized the order across participants to control for the effects of order. However, given that we did find systematic shifts in the CSPC effect for unbiased items, we can also ask whether the shift between phases was asymmetrical.

@@

@@

## Design 
Experiment 1 used a 2x3 within-subjects design with proportion congruent (25%, 50%, and 75%) and  congruency (congruent and incongruent) as factors. The experiment consisted of 288 flanker trials (96 per proportion congruent condition) and did not include any counting trials or instructions informing participants about the proportion manipulations. The design in every other respect was identical to Experiment 1.

@@

The questions of interest were whether participants would learn the associations between the context cues without the task-relevance manipulation (i.e., the counting task) and whether responding to the unbiased cues would be influenced by the presence of the biased cues. To that end, mean correct RTs and mean error rates were submitted to an analysis of variance (ANOVA) with proportion congruent (25%, 50%, and 75%) and congruency (congruent and incongruent) as within-subject factors.

The results of the RT analysis revealed the critical, two-way interaction between proportion congruency and congruency to be non-significant, `r RT_ANOVA_E2$PC_Congruency`. Similarly, the corresponding error analysis also resulted in a non-significant two-way interaction between the task-relevant context and unbiased-item congruency, `r ACC_ANOVA_E2$PC_Congruency`. Therefore, there was no evidence that the context cues, absent the counting task, systematically influenced the congruency effect.

@@

@@

## Design 
Experiment 1 used a 2x3 within-subjects design with proportion congruent (25%, 50%, and 75%) and  congruency (congruent and incongruent) as factors. The experiment consisted of 288 flanker trials (96 per proportion congruent condition) and did not include any counting trials or instructions informing participants about the proportion manipulations. The design in every other respect was identical to Experiment 1.

@@

@@

# General Discussion
The aim of the current study was to determine whether task-relevance plays a role in the contextual recruitment of selective attention. Specifically, we tested whether manipulating the relative task-relevance of context cues could cause participants to ignore a previously learned context-association and apply a new association. We created a frequency unbiased context cue that shared features with two frequency biased contexts and used a feature-counting task to manipulate the task-relevance of context dimensions across two blocks of trials. Critically, halfway through the experiment participants received new instructions changing the task-relevant feature from one frequency biased cue to the other.

>>>>

Real-world environments can offer a plethora of context cues. Consider someone writing in a coffee shop. The coffee shop is a complex environment with numerous context cues (e.g., pictures on the wall, chairs, coffee, laptop), each of which we have different experiences with and afford different learned associations. Although a laptop might cue attentional processing helpful for writing, other features of the coffee shop might cue attentional processing related to other experiences like socializing and people-watching. Given that all of these cues are available to set attentional priorities, what determines which associative relationships will be used to guide attention? On the one hand, context-based retrieval of attentional priorities might be completely obligatory: the mere presence of learned cues might trigger attention in a ballistic manner [e.g., @schneider_controlled_1977; @shiffrin_controlled_1977]. A writer's attention in this case would be at the mercy of the environment. On the other hand, our current task-goals might constrain the expression of previously learned associations. For instance, the writer's current goals may allow the laptop to take priority over other environmental cues that are irrelevant to the task at hand.

Historically, perspectives of automaticity and attentional control have emphasized the role of practice and likened the learning process to the development of automaticity in skill-learning [@schneider_controlled_1977; @shiffrin_controlled_1977; @logan_toward_1988; @logan_attention_1992]. Under this view, forming an association requires many repeated experiences pairing a cue with a set of attentional priorities. Once formed however, it becomes habitual or hard-wired and the mere presence of the cue is sufficient to automatically adjust attention. Learned associations are therefore ballistic, difficult to suppress, and difficult to unlearn [e.g., @schneider_controlled_1977; @shiffrin_controlled_1977; @posner_attention_1975; @posner_facilitation_1975; @treisman_contextual_1960; @treisman_selective_1964; @treisman_strategies_1969]. This view is reflected in modern work on contextual cuing where it is assumed that learned associations require practice [e.g., @lehle_fly_2008; @crump_learning_2016], form without awareness or intention, and are triggered automatically by the mere presence of the cue [@crump_context-specific_2006; @crump_contextual_2010; @crump_context-specific_2008; @king_priming_2012; @reuss_unconscious_2014; though, see @brosowsky_context-specific_2016].

Memory-based perspectives however, take a more nuanced view [@norman_memory_1969; @norman_memory_1976; @abrahamse_grounding_2016; @brosowsky_memory-guided_2018; @crump_learning_2016; @egner_creatures_2014]. Here, practice is only important to the extent that it improves memory retrieval. Learning an association is the process of encoding experiences where attentional priorities and cues co-occur. Memory retrieval is a similarity-based, cue-driven process [e.g., @jacoby_interpreting_1978; @hintzman_minerva_1984; @hintzman_schema_1986; @hintzman_judgments_1988]. Presumably, the features of our immediate environment automatically cue memory and reinstate the retrieved attentional priorites [@logan_toward_1988; @logan_attention_1992]. If a few prior experiences are easily retrievable (i.e., memories are distinct or very similar to the retrieval cue), they could support the reinstatement of a learned attentional association despite very little practice [e.g., @brosowsky_memory-guided_2018]. Conversely, if prior experiences are difficult to retrieve (i.e., memories are indistinct or low in similarity to the retrieval cue), then learned associations may not be expressed, even after much practice. To the extent that you could manipulate memory retrieval, you should be able manipulate the expression of learned associations.

@@

Recent work lends some credence to the possibility that task-relevance is necessary for establishing and using context-attention associations. Crump, Vaquero, & Milliken [-@crump_context-specific_2008] used a Stroop prime-probe task where the distractor word is presented first followed by a color patch probe. Critically, they used the shape of the color patch probe, either a square or a circle, as the contextual dimension; one associated with a high frequency of conflict, the other with a low frequency. In one experiment participants were made aware of the shape-proportion contingencies but given no additional instructions. In the second, participants were given a secondary task to count the number of trials that contained a square. The first experiment failed to find evidence for context-specificity, whereas the second did. Although not testing task-relevance directly, @canadas_social_2013 found similar effects. In this study they used images of male and female faces as context cues in a flanker task. When given instructions to think of the faces as members of the gender categories they found CSPC effects, but when given instructions to think of the faces as individuals they did not. These results suggest that making a context dimension relevant to the on-going task is important for producing CSPC effects.

<!-- Furthermore, most commonly used context cue is location [@brosowsky_context-specific_2016; @corballis_independent_2003; @crump_context-specific_2006; @crump_context-specific_2008; @crump_learning_2016; @weidler_attentional_2018; @weidler_transfer_2016], often claimed to be irrelevant to the on-going conflict task. Although in some sense, the location is irrelevant for identifying the color of a word in a Stroop task, it seems disingenuous to consider it completely task-irrelevant. In order to identify the target, the participant must first identify and orient to its location on the screen. In this is sense, it is relevant to their ability to complete the task. Further, there is some evidence that location information may receive priority during encoding [@logan_what_1998; @mayr_spatial_1996]. Similar arguments could be made about other context features that are inherent to processing the target stimulus, like font-type [@bugg_multiple_2008] and color [@vietze_context_2009].  -->

@@

Previous work has shown that shifting between list-wide proportion congruent phases produced asymmetrical effects. @abrahamse_attention_2013 found that shifting from a mostly-congruent to mostly-incongruent list produced a large change in the size of the congruency effect, whereas shifting from a mostly-incongruent to mostly-congruent list produced a relatively small change in congruency effects. The authors posited a learned attentional modulation in the mostly-incongruent block had carried over into the mostly-congruent block reducing what, typically, would be a large congruency effect. In our study, we were primarily interested in whether (a) participants could use context cues despite the presence of competing, contradictory cues, and (b) whether participants could switch from the use of one context cue to another. We therefore randomized the order across participants to control for the effects of order. However, given that we did find systematic shifts in the CSPC effect for unbiased items, we can also ask whether the shift between phases was asymmetrical.

@@

We made two assumptions in the design of Experiment 1: First, we assumed that task-relevance was required for participants to learn about the associations between context cues and the biased proportion congruency manipulations. Our task was inspired by @crump_flexibility_2009 who failed find context-specific effects using a task-irrelevant shape cue until they used a counting task to make the shapes relevant. However, location is the most-often used context cue in these kinds of paradigms (e.g., ). In these cases, no task-relevance manipulation is necessary to produce CSPC effects using location cues. Though, it is arguable whether location is truly task-irrelevant when localizing the target is necessary for responding. Regardless, it is unclear in our paradigm whether participants required the counting task to learn about the biased proportion congruency manipulations.

Second, we assumed that the biased proportion congruency manipulations would have no systematic effect on the unbiased items without the counting task. Although we randomized all aspects of the task across participants to control for potential confounds, it still remains unclear whether the biased items would influence how paricipants responded to the unbiased items.

Experiment 2, therefore, serves as a control experiment to test these two assumptions. Experiment 2 is identical to Experiment 1 with two important changes: First, we removed the counting task and gave no explicit instructions about how to use the context cues; We only informed participants that images would be present throughout the task. This allowed us to determine if there was any systematic effect of biased items on the unbiased items absent the counting task. Second, we altered the proportion congruency of the biased item sets from 100%/0% to 75%/25%. This allowed us to measure the congruency effect for the biased item sets to determine whether participants learned about the associations between proportion congruency and the context cues without the counting task [e.g., @crump_flexibility_2009]. If this were the case, we would expect larger congruency effects for the 75% proportion congruent items as compared to the 25% proportion congruent items.

@@

## Design 
Experiment 2 used a 2x3 within-subjects design with proportion congruent (25%, 50%, and 75%) and  congruency (congruent and incongruent) as factors. The experiment consisted of 288 flanker trials (96 per proportion congruent condition) and did not include any counting trials or instructions informing participants about the proportion manipulations. The design in every other respect was identical to Experiment 1.

@@

The questions of interest were whether participants would learn the associations between the context cues without the task-relevance manipulation (i.e., the counting task) and whether responding to the unbiased cues would be influenced by the presence of the biased cues. To that end, congruency effects (incongruent minus congruent performance) for RTs and error rates were submitted to an analysis of variance (ANOVA) with proportion congruency (25%, 50%, and 75%) as the sole within-subject factor.

The results of the RT analysis revealed the effect of proportion congruency to be non-significant, `r RT_ANOVA_E2$PC`, with moderate evidence in favor of the null, `r RT_BF_E2[[1]]`. Similarly, the corresponding error analysis also resulted in a non-significant effect of proportion congruency, `r ACC_ANOVA_E2$PC`, with strong evidence in favor of the null, `r ACC_BF_E2[[1]]`. Therefore, there was no evidence that the context cues, absent the counting task, systematically influenced the congruency effects in any way.

@@

In the current study, we have been homing in on the importance of task-relevance for the contextual recruitment of selective attention. In Experiment 1, we found that task-relevance allowed participants to exploit context cues in the presence of a competing, contradictory cue and subsequently ignore the previously learned association in favor of a different context cue. In Experiment 2, we conceptually replicated previous work showing that task-relevance was necessary for participants to learn about the biased proportion congruency associations [e.g., @crump_flexibility_2009] and ruled out possible confounding effects of the biased items.

@@

## Design 
Experiment 3 used a 2x3 mixed design with task-relevant context (0% and 100%) and unbiased-item congruency (congruent and incongruent) as the within-subject factors. The experiment consisted of 288 flanker trials (96 per proportion congruent condition) and did not include any counting trials. However, participants were explicitly informed about the proportion congruency manipulations and instructed to make used of the context cues (see the Procedure section below for more details). Therefore, unlike Experiment 1, context images were made task-relevant through explicit instructions rather than the counting task. The design in every other respect was identical to Experiment 1.

@@

Importantly, participants were given further instructions to exploit the proportion congruency associated with context cues.

@@

# General Discussion
The aim of the current study was to determine whether task-relevance plays a role in the contextual recruitment of selective attention. In Experiment 1, we tested whether manipulating the relative task-relevance of context cues could cause participants to ignore a previously learned context-association and apply a new association. We created a frequency unbiased context cue that shared features with two frequency biased contexts and used a feature-counting task to manipulate the task-relevance of context dimensions across two blocks of trials. Critically, halfway through the experiment participants received new instructions changing the task-relevant feature from one frequency biased cue to the other.
